Sample Clustering

### 2017-02-28 Clustering.Rmd
Updating the clustering analysis to generate full report

---
title: "Sample Clustering"
author: "Sarah Killcoyne"
date: "23/02/2016"
output: html_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(GGally)
library(plyr)
library(xlsx)
library(dplyr)
library(pander)
library(Hmisc)
library(ggdendro)
library(reshape2)
library(GenomicRanges)
library(gridExtra)
library(biomaRt)
library(multtest)
library(tibble)
library(mclust)

@@

```{r global_pt, echo=F, message=F, warning=F}
be = subset(patient.info, Pathology == 'BE')
lgd = subset(patient.info, Pathology <= 'LGD')

@@

set.seed(99)
mtp1 = MTP(X=t(patient.info[,c('Barretts.Cellularity','Total.Reads')]), Y=as.integer((as.integer(patient.info$Pathology) > 3)), robust=T, get.adjp=T, na.rm=T)
mtp2 = MTP(X=t(patient.info[,c('Barretts.Cellularity','Total.Reads')]), Y=patient.info$Status, robust=T, get.adjp=T, na.rm=T)
```

Currently with multi test correction the samples for which cellularity and #of reads are available show no difference for either status (P vs NP)
`r pander(summary(mtp1), justify='left')`

or for samples that are LGD/IMC/HGD vs BE
`r pander(summary(mtp1), justify='left')`

@@

```{r pca, echo=F, warning=F, message=F, fig.width=20, fig.height=40}
sum.patient.data = arrange(sum.patient.data, Status, total.samples)

@@

    eigenvalues = pca$sdev^2
    screeplot(pca, main=paste(pt, sv, 'scree'))

@@

pplots = lapply(subset(sum.patient.data, Status == 'P')$Patient, plot.pca)
grid.arrange(grobs=pplots, nrow=length(subset(sum.patient.data, Status == 'P')$Patient), top="Progressors")

npplots = lapply(subset(sum.patient.data, Status == 'NP')$Patient, plot.pca)
grid.arrange(grobs=npplots, nrow=length(subset(sum.patient.data, Status == 'NP')$Patient), top="Non-Progressors")

@@

I tried were several mixture models (e.g. EM, kmeans) but wasn't able to find a pattern that could reliably separate the few mixtures I did see

@@

#### EM
```{r clust, echo=T, message=T, warning=F, eval=T, fig.height=4, fig.width=4, eval=T}

em.dr<-function(patient, status, x1, var.regions) {
  clusters=c(2:4)
  #llk = matrix(ncol=length(clusters),nrow=nrow(sum.patient.data), dimnames=list(sum.patient.data$Patient, clusters))

  llk = vector(length=length(clusters), mode='numeric')
  names(llk) = as.character(clusters)

## Ok, this doesn't separate regions well either normalized or raw segment values.  Typically the clusters are highly overlapped and not very descriptive
#for (pt in sum.patient.data$Patient) {
    print(pt)
#    status = subset(sum.patient.data, Patient == pt)$Status
#    print(ncol(patient.data[[pt]]$seg.vals[,-(1:5)]))
#    x1 = as.matrix(patient.data[[pt]]$seg.vals[,-(1:5)])
#    x1 = x1[,patient.data[[pt]]$info$Samplename]
    #x1 = as.matrix(patient.data[[pt]]$norm.seg.vals)

    for (g in clusters) {
      print(g)
      model = densityMclust(x1, G=g)
      highCluster = which.max(colSums(model$parameters$mean))

@@

      a1 = rownames(x1[which(model$classification == highCluster),] )
      if (nrow(var.regions) > 0) {
        am = merge(a1, var.regions, all=T)
        pander( as.data.frame(rownames(x1[which(model$classification == highCluster),])),
             caption=paste("From the cluster with the highest means the regions above are selected, of these ",length(setdiff(rownames(var.regions), a1)),
                           " regions are not found in the variable regions selected using cutoffs.", length(intersect(rownames(var.regions), a1)), " are in both."))
      } else {
        pander(rownames(x1[which(model$classification == highCluster),]),
             caption=paste("From the cluster with the highest means the regions above are selected. This patient had no variable regions selected based on cutoffs"))

      }

      llk[as.character(g)] = model$loglik
    }
    return(llk)
}

llk = as.data.frame(matrix(nrow=nrow(sum.patient.data),ncol=3,dimnames=list(c(sum.patient.data$Patient), c(2:4))))
for (pt in sum.patient.data$Patient) {
    print(pt)
    status = subset(sum.patient.data, Patient == pt)$Status

@@

    llk[pt,] =  em.dr(pt,status,as.matrix(x1), var[,-(1:5)])
  }

ggplot(melt(as.matrix(llk)), aes(x=Var1, y=value, color=Var1)) + geom_point() + 
  theme(axis.text.x=element_text(angle = 45), legend.position='none')
  labs(title="Raw segment value log liklihoods", x='', y='loglik')

# Log liklihood doesn't separate them individually or the range
llk.diff = apply(llk, 1, function(x) diff(range(x)) )
wilcox.test(llk.diff[subset(sum.patient.data, Status == 'NP')$Patient], llk.diff[subset(sum.patient.data, Status == 'P')$Patient])

@@

#### Kmeans
```{r kmeansraw, echo=F, message=F, warning=F, eval=F, fig.height=6, fig.width=6} 
## Kmeans
#   x1 = as.matrix(patient.data[[pt]]$seg.vals[,-(1:5)])
# 
for (pt in sum.patient.data$Patient) {
    print(pt)
    status = subset(sum.patient.data, Patient == pt)$Status
    x1 = patient.data[[pt]]$seg.vals
    rownames(x1) = apply(patient.data[[pt]]$seg.vals[,c(1:4)], 1, function(x) strip.whitespace(paste(x[['chrom']], ':', x[['start.pos']], '-', x[['end.pos']], sep=''))) 
    x1 = x1[,patient.data[[pt]]$info$Samplename]

    fit = kmeans(x1, 3)
    autoplot(fit, data=x1, label=T) + labs(title=paste(pt, '(', status, ') raw seg', sep=''))
    pander(patient.data[[pt]]$variable.regions[,(1:4)], caption='Variable regions selected by cutoffs')
}
```

### Normalized segmented values
```{r clust2, echo=T, message=T, warning=F, eval=T, fig.height=4, fig.width=4, eval=T}
llk = as.data.frame(matrix(nrow=nrow(sum.patient.data),ncol=3,dimnames=list(c(sum.patient.data$Patient), c(2:4))))
for (pt in sum.patient.data$Patient) {
    print(pt)
    status = subset(sum.patient.data, Patient == pt)$Status
    x1 = patient.data[[pt]]$norm.seg.vals
    rownames(x1) = apply(patient.data[[pt]]$seg.vals[,c(1:4)], 1, function(x) strip.whitespace(paste(x[['chrom']], ':', x[['start.pos']], '-', x[['end.pos']], sep=''))) 
    x1 = x1[,patient.data[[pt]]$info$Samplename]

@@

    llk[pt,] = em.dr(pt,status,x1, var)
  }

ggplot(melt(as.matrix(llk)), aes(x=Var1, y=value, color=Var1)) + geom_point() + 
  theme(axis.text.x=element_text(angle = 45), legend.position='none') + 
  labs(title="Normalized segment value log liklihoods", x='', y='loglik')

# Log liklihood doesn't separate them individually or the range
llk.diff = apply(llk, 1, function(x) diff(range(x)) )
wilcox.test(llk.diff[subset(sum.patient.data, Status == 'NP')$Patient], llk.diff[subset(sum.patient.data, Status == 'P')$Patient])

@@

#### Kmeans
```{r kmeansnorm, echo=F, message=F, warning=F, eval=F, fig.height=6, fig.width=6} 
for (pt in sum.patient.data$Patient) {
    print(pt)
    status = subset(sum.patient.data, Patient == pt)$Status
    x1 = patient.data[[pt]]$norm.seg.vals
    rownames(x1) = apply(patient.data[[pt]]$seg.vals[,c(1:4)], 1, function(x) strip.whitespace(paste(x[['chrom']], ':', x[['start.pos']], '-', x[['end.pos']], sep=''))) 
    x1 = x1[,patient.data[[pt]]$info$Samplename]

    fit = kmeans(x1, 3)
    print(autoplot(fit, data=x1, label=T) + labs(title=paste(pt, '(', status, ') normalized seg', sep='')))
    pander(patient.data[[pt]]$variable.regions[,(1:4)], caption='Variable regions selected by cutoffs')
}
```

@@

```{r clustN, echo=T, message=T, warning=F, eval=F}
## Normalized Segmentation Values
#Doesn't work for the normalized values.

for (pt in subset(sum.patient.data, Status == 'P')$Patient) {
  if (pt == 'AH0254') next
  print(pt)
  print(paste('norm.seg.vals cols:', ncol(patient.data[[pt]]$norm.seg.vals)))

  x1 = patient.data[[pt]]$norm.seg.vals
  model = Mclust(t(x1))
  print(paste('loglik', round(model$loglik,2)))

  tryCatch({
    mod1dr = MclustDR(model, lambda = 1)
    #summary(mod1dr)
    table(mod1dr$mixcomp)

    if (length(table(mod1dr$mixcomp)) > 1) {
      plot(mod1dr, what = "evalues")
      title(main=pt)
      #plot(mod1dr, what = "density")
    }
  })
}

So tried AP as that doesn't require the pre-definition for number of clusters.  It appears to give good separation, though there's a suggestion of two different types of progressors.

### Affinity Propagagion clustering described by Frey & Dueck, 2007

First on the raw segmentation values

```{r apclust, echo=T, message=T, warning=F, fig.height=10, fig.width=10, fig.align='left'}
q=0.2

@@

  last = ncol(x1)
  if (patient.data[[pt]]$info$Status[1] == 'P' & pt != 'AD0252')
    last = sort(grep('HGD|IMC', patient.data[[pt]]$info$Pathology), decreasing=T)[1]
  # if (ncol(x1) <= 15) {
  #   plot(ac, x1[,c(c(1:ncol(x1))[-last], last)], main=pt)
  # } else {
  #   plot(ac, x1[,c(1, sort(sample(c(1:(ncol(x1)-1))[c(-1,-last)], 8)) ,last) ], main=pt)
  # }

@@

  if (nrow(tmp) <= 0) {
    sim[pt,c('gains', 'loss')] = c(0,0)
  } else {
    sim[pt,'gains'] = length(tmp[tmp>=1.1])
    sim[pt,'loss'] =  length(tmp[tmp<=0.9])
  }
}
sim$id = rownames(sim)

@@

## Number of variable regions isn't so good, but maybe...
sim %>%
  group_by(Status) %>%
  summarise( mean = mean(n.var), sd = sd(n.var), med = median(n.var))
t.test( subset(sim, Status == 'NP')$n.var, subset(sim, Status == 'P')$n.var )$p.value

@@

g2 = ggplot(sim[,c('Status','clusters')], aes(y=clusters, x=Status, fill=Status)) + geom_boxplot() + 
  geom_jitter(width=0.2) + stat_boxplot(geom ='errorbar') +
  labs(title=paste("p-value", signif(ttpc, 3)), y='num. clusters')

grid.arrange(g1, g2, top='Raw seg values')

# fit = glm(Status~clusters, sim, family=binomial(link='logit'))
# pander(fit, justify='left')
# 
# sim = sim[, c('Status', 'clusters', 'ns','ss')]
# actualMatrix = matrix(0,1000,2,dimnames=list(c(), c('prec','recall')))
# for (i in 1:nrow(actualMatrix)) {
#   trainingRows = sample(1:nrow(sim), nrow(sim)-5)
#   train = sim[trainingRows,]
#   test = sim[-trainingRows,]
#   
#   fit = glm( (as.integer(Status)-1)~clusters, train, family=binomial(link='logit'))
#   pred = predict(fit, test, se.fit=F) # probabilities -- of which outcome though?
#   
#   actual = test$Status == 'P'
#   predicted = pred>0
#   
#   actualMatrix[i,'prec'] = getPrecision( actual, predicted )
#   actualMatrix[i, 'recall'] = getRecall( actual, predicted )
# }
# 
# ggplot(as.data.frame(actualMatrix), aes(x=recall, y=prec)) + geom_point() + 
#   geom_hline(yintercept = 0.5, col='red', alpha=0.5) +
#   labs(title='Precision/Recall for Progressor/NP prediction')

rc = cor.test(sim$max.sd, sim$clusters)
```

@@

We needed to check if cellularity was driving the difference in clusters. Overall that does not appear to be the case if we use the maximum standard deviation across all samples in a patient as a proxy: r=`r rc$estimate`, p-value=`r signif(rc$p.value, 2)`

### Are the clusters random?

```{r example_pt}
pt = 'PR1_HIN_042'
patient.data[[pt]]$info = arrange(patient.data[[pt]]$info, Endoscopy.Year, desc(Pathology))

@@

@@

hgmx = as.data.frame(hgmx)

ggplot(subset(hgmx, x > 0), aes(x=chr,y=P.x,color=factor(cluster))) + ylim(0,1)+ 
  geom_point() + geom_line(alpha=0.1)

ggplot(subset(hgmx, x > 0 & cluster == 9), aes(x=factor(chr),y=P.x,fill=factor(cluster))) + ylim(0,0.5 )+ 
  geom_bar(colour='grey', stat='identity', position=position_dodge())

So what about using the normalized segmented values?

```{r apclustN, echo=T, message=T, warning=F, fig.height=8, fig.align='left'}
# Should this be performed on the normalized seg values instead?
sim3 = sum.patient.data[,c(2:4)]
rownames(sim3) = sum.patient.data[,1]
for (pt in sum.patient.data$Patient) {
  x1 = patient.data[[pt]]$norm.seg.vals
  ac = apcluster(negDistMat(x1, r=2), q=q, details=T, includeSim=T)
  #plot(ac)

@@

ggplot(sim3[,c('Status','ss')], aes(y=ss, x=Status, fill=Status)) + geom_boxplot() + geom_jitter(width=0.2) + 
  labs(title=paste("Norm seg values p-value", signif(ttp, 3)), y='sum similarity')
```

@@

Patients first

```{r xval_pt, echo=T, warning=F, message=T, fig.align='left', fig.height=4, eval=F}
# Leave out patients
loo_xval = matrix(ncol=3, nrow=0, dimnames=list(c(), c('net.similarity', 'sum.similarity', 'clusters')))
for (i in 1:nrow(sum.patient.data)) {
  simL = sum.patient.data[-i,c(2:4)]
  rownames(simL) = sum.patient.data[-i,1]

@@

Then randomly leave out samples

```{r xval_smp, echo=F, warning=F, message=T, fig.align='left', fig.height=6, eval=F}

# Leave out samples?
looS_xval = matrix(ncol=3, nrow=0, dimnames=list(c(), c('net.similarity', 'sum.similarity', 'clusters')))
for (i in 1:20) { 
  simX = sum.patient.data[,c('Status','years')]
  rownames(simX) = sum.patient.data[,'Patient']

@@

## (xval) Re run clusters on the Progressors without the HGD sample
There is still a difference between the progressors and non, though the p-value does increase. Some of the progressors do start looking more like NPs in their similarity measure but not all of them.

```{r nofinal, echo=F, message=F, warning=F, fig.height=6, fig.width=10}

@@

for (pt in sum.patient.data$Patient) {
  print(pt)
  x1 = apclust.data(patient.data[[pt]]$seg.vals, patient.data[[pt]]$info$Samplename)
  if (subset(sum.patient.data, Patient == pt)$Status == 'P' & length(which(grepl('HGD|IMC', patient.data[[pt]]$info$Pathology))) > 0) {
    final = grep('HGD|IMC', patient.data[[pt]]$info$Pathology)
    x1 = apclust.data(patient.data[[pt]]$seg.vals, patient.data[[pt]]$info$Samplename[-final])
  }
  ac = apcluster(negDistMat(x1, r=2), q=q)

@@

t.test(subset(simX, Status == 'NP')$n.var, subset(simX, Status == 'P')$n.var)

#pander(subset(simX, Status == 'P' & ss <= mean(ss)), caption="These patients are still strongly in the Progressor group")

pander(subset(simX, Status == 'P' & clusters >= mean(clusters)), caption="These patients are more similar to Non-Progressors without their HGD sample(s)")

fit = glm(Status~clusters, simX, family=binomial(link='logit') )
summary(fit)

@@

## And why q=0.2
```{r whyq, echo=F, message=F, warning=F}
sum.patient.data = arrange(sum.patient.data, Status)

pvals = as.data.frame(matrix(ncol=3,nrow=0,dimnames=list(c(), c('p.clusters','mean.clusters', 'sd.clusters'))))
status.clusters = list()
for (qv in seq(0,0.9,0.1)) {
  print(qv)

  ## Using the segmentation values
  simPC = sum.patient.data[,c(2:4)]
  rownames(simPC) = sum.patient.data[,1]
  for (pt in sum.patient.data$Patient) {
    patient.data[[pt]]$info = arrange(patient.data[[pt]]$info, Endoscopy.Year, desc(Pathology))

@@

pander(clus[,c('qv','NP','P')], caption="Mean of clusters by patient status per q parameter.")

ggplot(melt(pvals, measure.vars=c('p.clusters'), id.vars='qv'),aes(x=factor(qv),y=value, color=variable)) + 
  geom_point() + 
  labs(title='P vs NP p-values per q parameters', y='pvalue NP vs P', x='q param')

>>>>

---
title: "Sample Clustering"
author: "Sarah Killcoyne"
date: "23/02/2016"
output: 
  html_document: 
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
library(ggplot2)
library(GGally)
library(plyr)
library(xlsx)
library(dplyr)
library(pander)
library(Hmisc)
library(ggdendro)
library(reshape2)
library(GenomicRanges)
library(gridExtra)
library(biomaRt)
library(multtest)
library(tibble)
library(mclust)
library(ggfortify)

@@

We have a few questions we needed to answer with this data.  The pairwise correlation plots appeared to show a difference between progressors and non-progressors as well as some differences between samples within a single patient.  So...

1. Within a single patient, can we quantify in some way the diversity that appears to be present in the pairwise correlation plots? 
2. Given a diversity measure within a single patient, can we differentiate the progressors from the non-progressors?  Can we do it prior to the HGD/cancer sample?
3. Is there then a way to select patient-specific thresholds for calling variable regions?  Alternately, given what we know about the cancer can we presume diploidy and treat these samples as if we had absolute copy number?

The first two questions were addressed with the clustering below.  The third is in progress currently.

```{r global_pt, echo=F, message=F, warning=F, error=F}
be = subset(patient.info, Pathology == 'BE')
lgd = subset(patient.info, Pathology <= 'LGD')

@@

set.seed(99)
mtp1 = MTP(X=t(patient.info[,c('Barretts.Cellularity','Total.Reads')]), Y=as.integer((as.integer(patient.info$Pathology) > 3)), 
           get.adjp=T, na.rm=T, test='t.twosamp.equalvar')
mtp2 = MTP(X=t(patient.info[,c('Barretts.Cellularity','Total.Reads')]), Y=patient.info$Status, 
            get.adjp=T, na.rm=T, test='t.twosamp.equalvar')
```

Currently with multi test correction the samples for which cellularity and #of reads are available show no difference for either status (P vs NP)
`r pander(mtp2@adjp, justify='left', caption="Adjusted p-value")`

or for samples that are LGD/IMC/HGD vs BE
`r pander(mtp1@adjp, justify='left', caption="Adjusted p-value")`

However, it should be noted that pathologists do not like calling cellularity on Barrett's generally so it's a fairly inaccurate measure.
`r pander(table( patient.info[,c('Status','Barretts.Cellularity')]), caption="Numbers of cellularity calls by status")`

@@

```{r pca, echo=F, warning=F, message=F, fig.width=8, fig.height=4}
sum.patient.data = arrange(sum.patient.data, Status, total.samples)

@@

    eigenvalues = pca$sdev^2
    #screeplot(pca, main=paste(pt, sv, 'scree'))

@@

pplots = lapply(subset(sum.patient.data, Status == 'P')$Patient, plot.pca)
npplots = lapply(subset(sum.patient.data, Status == 'NP')$Patient, plot.pca)
```

```{r echo=F, message=F, warning=F, pcaplots, fig.align='left', fig.height=30, fig.width=20}
grid.arrange(grobs=pplots, nrow=length(subset(sum.patient.data, Status == 'P')$Patient), top="Progressors")
grid.arrange(grobs=npplots, nrow=length(subset(sum.patient.data, Status == 'NP')$Patient), top="Non-Progressors")
```

# Clustering

@@

I tried were several mixture models (e.g. EM, kmeans) but wasn't able to find a pattern that could reliably separate the distributions.

@@

First using EM (mclust) looking at three different values for G (2,3,4).

#### EM
```{r clust, echo=F, message=F, warning=F, fig.height=4, fig.width=4}

em.dr<-function(patient, status, x1, var.regions) {
  cat(paste('**', patient, '**'))
  clusters=c(2:4)

  llk = list()

  for (g in clusters) {
    model = densityMclust(x1, G=g)
    highCluster = which.max(colSums(model$parameters$mean))

@@

    # a1 = rownames(x1[which(model$classification == highCluster),] )
    # if (nrow(var.regions) > 0) {
    #   am = merge(a1, var.regions, all=T)
    #   pander( as.data.frame(rownames(x1[which(model$classification == highCluster),])), caption=paste("From the cluster with the highest means the regions above are selected, of these ",length(setdiff(rownames(var.regions), a1))," regions are not found in the variable regions selected using cutoffs.", length(intersect(rownames(var.regions), a1)), " are in both."))
    # } else {
    #   pander(as.data.frame(a1),caption=paste("From the cluster with the highest means the regions above are selected. This patient had no variable regions selected based on cutoffs"))
    # 
    # }
    llk[[as.character(g)]] = model
  }
  return(llk)
}

llk = as.data.frame(matrix(nrow=nrow(sum.patient.data),ncol=3,dimnames=list(c(sum.patient.data$Patient), c(2:4))))
for (pt in sum.patient.data$Patient) {
    status = subset(sum.patient.data, Patient == pt)$Status

@@

    models = em.dr(pt,status,as.matrix(x1), var[,-(1:5)])
    llk[pt,] = sapply(models, function(x) x$loglik)
  }

ggplot(melt(as.matrix(llk)), aes(x=factor(Var1), y=value, color=factor(Var2))) + 
  geom_point() +
  theme(axis.text.x=element_text(angle = 45)) + 
  labs(title="Raw segment value log liklihoods", x='', y='loglik')

# Log liklihood doesn't separate them individually or the range
llk.diff = apply(llk, 1, function(x) diff(range(x)) )
#t.test(llk.diff[subset(sum.patient.data, Status == 'NP')$Patient], llk.diff[subset(sum.patient.data, Status == 'P')$Patient])

@@

#### Kmeans

```{r kmeansraw, echo=F, message=F, warning=F, eval=F, fig.height=6, fig.width=6} 
for (pt in sum.patient.data$Patient) {
    status = subset(sum.patient.data, Patient == pt)$Status
    x1 = patient.data[[pt]]$seg.vals
    rownames(x1) = apply(patient.data[[pt]]$seg.vals[,c(1:4)], 1, function(x) strip.whitespace(paste(x[['chrom']], ':', x[['start.pos']], '-', x[['end.pos']], sep=''))) 
    x1 = x1[,patient.data[[pt]]$info$Samplename]

    fit = kmeans(x1, 3)
    autoplot(fit, data=x1, label=F) + labs(title=paste(pt, '(', status, ') raw seg', sep=''))
    pander(patient.data[[pt]]$variable.regions[,(1:4)], caption='Variable regions selected by cutoffs')
}
```

### Normalized segmented values

#### EM

Row-normalized (per region)

```{r clust2, echo=F, message=F, warning=F, fig.height=4, fig.width=4}
llk = as.data.frame(matrix(nrow=nrow(sum.patient.data),ncol=3,dimnames=list(c(sum.patient.data$Patient), c(2:4))))
for (pt in sum.patient.data$Patient) {
    status = subset(sum.patient.data, Patient == pt)$Status
    x1 = patient.data[[pt]]$norm.seg.vals
    rownames(x1) = apply(patient.data[[pt]]$seg.vals[,c(1:4)], 1, function(x) strip.whitespace(paste(x[['chrom']], ':', x[['start.pos']], '-', x[['end.pos']], sep=''))) 
    x1 = x1[,patient.data[[pt]]$info$Samplename]

@@

        models = em.dr(pt,status,as.matrix(x1), var[,-(1:5)])
    llk[pt,] = sapply(models, function(x) x$loglik)
  }

ggplot(melt(as.matrix(llk)), aes(x=Var1, y=value, color=factor(Var2))) + geom_point() + 
  theme(axis.text.x=element_text(angle = 45)) + 
  labs(title="Normalized segment value log liklihoods", x='', y='loglik')

# Log liklihood doesn't separate them individually or the range
llk.diff = apply(llk, 1, function(x) diff(range(x)) )
#wilcox.test(llk.diff[subset(sum.patient.data, Status == 'NP')$Patient], llk.diff[subset(sum.patient.data, Status == 'P')$Patient])

@@

#### Kmeans
```{r kmeansnorm, echo=F, message=F, warning=F, fig.height=6, fig.width=6} 
for (pt in sum.patient.data$Patient) {
    status = subset(sum.patient.data, Patient == pt)$Status
    x1 = patient.data[[pt]]$norm.seg.vals
    rownames(x1) = apply(patient.data[[pt]]$seg.vals[,c(1:4)], 1, function(x) strip.whitespace(paste(x[['chrom']], ':', x[['start.pos']], '-', x[['end.pos']], sep=''))) 
    x1 = x1[,patient.data[[pt]]$info$Samplename]

    fit = kmeans(x1, 3)
    print(autoplot(fit, data=x1, label=F) + labs(title=paste(pt, '(', status, ') normalized seg', sep='')))
    pander(patient.data[[pt]]$variable.regions[,(1:4)], caption='Variable regions selected by cutoffs')
}
```

In neither the raw or normalized matrices could I reliably distinguish clusters from each other within a single patient. Some patients separate well (AH0254) others don't separate at all.  I looked as well for a high level measuremeant that might separate the patients by progressor/non. Log liklihoods did not show any difference between patients in the mixture model, and with poor clusters separation I was a bit stumped as to what else might work.

@@

I then tried Affinity Propagation (described by Frey & Dueck, 2007) as that doesn't require the pre-definition for number of clusters. My hypothesis was that the progressors would show a higher number of clusters, and that the variable regions might tend to show up in the same clusters.

### Raw segmentation values

```{r apclust, echo=F, message=F, warning=F, fig.height=10, fig.width=10, fig.align='left'}
q=0.2

@@

  last = ncol(x1)
  if (patient.data[[pt]]$info$Status[1] == 'P' & pt != 'AD0252')
    last = sort(grep('HGD|IMC', patient.data[[pt]]$info$Pathology), decreasing=T)[1]
  if (ncol(x1) <= 15) {
    plot(ac, x1[,c(c(1:ncol(x1))[-last], last)], main=pt)
  } else {
    plot(ac, x1[,c(1, sort(sample(c(1:(ncol(x1)-1))[c(-1,-last)], 8)) ,last) ], main=pt)
  }

@@

  if (nrow(tmp) <= 0) {
    sim[pt,c('gains', 'loss')] = c(0,0)
  } else {
    sim[pt,'gains'] = length(tmp[tmp>=1.1])
    sim[pt,'loss'] =  length(tmp[tmp<=0.9])
  }
}
sim$id = rownames(sim)
rc = cor.test(sim$max.sd, sim$clusters)

@@

## Number of variable regions isn't so good, but maybe...
#sim %>%
#  group_by(Status) %>%
#  summarise( mean = mean(n.var), sd = sd(n.var), med = median(n.var))
#t.test( subset(sim, Status == 'NP')$n.var, subset(sim, Status == 'P')$n.var )$p.value

@@

g2 = ggplot(sim[,c('Status','clusters')], aes(y=clusters, x=Status, fill=Status)) + geom_boxplot() + 
  geom_jitter(width=0.2) + stat_boxplot(geom ='errorbar') +
  labs(title=paste("p-value", signif(ttpc, 3)), y='num. clusters')
```

This plot compares the two possible measures of overall complexity from AP. The summed similarity value (top) and the number of clusters (bottom). In this case, the number of clusters appears to be a more robust measure.

```{r echo=F, message=F, warning=F, fig.width=5, fig.height=5}
grid.arrange(g1, g2, top='Raw seg values')
```

Based on the number of clusters in each patient, it appears that there are a higher number of clusters being selected by AP in the progressors vs the non, suggesting it may be able to relate it to diversity in the progressors CN. That metric appears to provide a good separation between the patients as well, though there's a suggestion of two different types of progressors.

```{r apglmclus, echo=F, message=F, warning=F}
fit = glm(Status~clusters, sim, family=binomial(link='logit'))
#pander(fit, justify='left')

sim = sim[, c('Status', 'clusters', 'ns','ss')]
actualMatrix = matrix(0,1000,2,dimnames=list(c(), c('prec','recall')))
for (i in 1:nrow(actualMatrix)) {
  trainingRows = sample(1:nrow(sim), nrow(sim)-5)
  train = sim[trainingRows,]
  test = sim[-trainingRows,]

  fit = glm( (as.integer(Status)-1)~clusters, train, family=binomial(link='logit'))
  pred = predict(fit, test, se.fit=F) # probabilities -- of which outcome though?

  actualMatrix[i,'prec'] = getPrecision( actual, predicted )
  actualMatrix[i, 'recall'] = getRecall( actual, predicted )
}

# ggplot(as.data.frame(actualMatrix), aes(x=recall, y=prec)) + geom_point() +
#   geom_hline(yintercept = 0.5, col='red', alpha=0.5) +
#   labs(title='Precision/Recall for Progressor/NP prediction')

@@

There is a question about whether the cellularity of the sample drives the differences in the clusters. Overall that does not appear to be the case if we use the maximum standard deviation across all samples in a patient as a proxy for cellularity measure: r=`r rc$estimate`, p-value=`r signif(rc$p.value, 2)`

#### Are the clusters random?

Especially with copy-number the clusters could simply be groups of neighboring regions from the same chromosomes as other analyses have shown high correlations between neighboring CN regions. So are the clusters that AP finds randomly distributed?

```{r example_pt, echo=F, warning=F, message=F, fig.width=10, fig.height=10}
pt = 'PR1_HIN_042'
patient.data[[pt]]$info = arrange(patient.data[[pt]]$info, Endoscopy.Year, Pathology)

@@

ac = apcluster(negDistMat(x1, r=2), q=q, details=T, includeSim=T)
plot(ac, x1, main=pt)

@@

hgmx = as.data.frame(hgmx)
```

```{r hgmx, echo=F, message=F, warning=F, fig.height=6, fig.width=6}
ggplot(subset(hgmx, x > 0), aes(x=chr,y=P.x,color=factor(cluster))) + ylim(0,0.5)+ 
  geom_point() + 
  geom_line(alpha=0.1) + 
  labs(title='Probability of cluster assignment', x='Chromosome', y='P(x)')
```

No cluster in any chromosome reached even a 50% probability, so it appears that the regions are reasonably randomly distributed across clusters rather than being made up of all of chromosome 1 or 17 etc.

```{r echo=F, warning=F, message=F}
ggplot(subset(hgmx, x > 0 & cluster == 9), aes(x=factor(chr),y=P.x,fill=factor(cluster),label=x)) + ylim(0,0.5) + 
  geom_bar(colour='grey', fill='purple', stat='identity', position=position_dodge()) + 
  geom_text(aes(y=P.x+0.005), position=position_dodge()) +
  labs(title='Cluster 9', x='P(x)', y='chromosome')
```

In the purple cluster are those unique regions to the HGD sample. So just as an example, the chromosomes that appear in that cluster are `r paste(subset(hgmx, cluster==9 & x>0)$chr, collapse=', ')` with a total of `r sum(subset(hgmx, cluster==9 & x>0)$x)` regions. The probabilities of each of these having shown up are:

`r pander( subset(hgmx, cluster==9 & x>0), caption="Cluster 9 probabilities are below 20%, or essentially random." )`

### Using the normalized values

So what about using the normalized segmented values for AP clustering instead?

```{r apclustN, echo=F, message=F, warning=F, fig.height=5, fig.width=5, fig.align='left'}
# Should this be performed on the normalized seg values instead?
sim3 = sum.patient.data[,c(2:4)]
rownames(sim3) = sum.patient.data[,1]
for (pt in sum.patient.data$Patient) {
  x1 = patient.data[[pt]]$norm.seg.vals
  ac = apcluster(negDistMat(x1, r=2), q=q, details=T, includeSim=T)
  #plot(ac)

@@

g1 = ggplot(sim3[,c('Status','ss')], aes(y=ss, x=Status, fill=Status)) + geom_boxplot() + geom_jitter(width=0.2) + 
  labs(title=paste("Norm seg values p-value", signif(ttp, 3)), y='sum similarity')

g2 = ggplot(sim3[,c('Status','clusters')], aes(y=clusters, x=Status, fill=Status)) + geom_boxplot() + geom_jitter(width=0.2) + 
  labs(title=paste("Norm seg values p-value", signif(ttp, 3)), y='n clusters')

grid.arrange(g1, g2, top='Normalized seg values')

The normalized values show no separation between progressor and non using either the summed similarity measure or the number of clusters. So for now we will use the raw segmentation value matrix.

### Cross-validation, LOO

@@

#### Patients first

```{r xval_pt, echo=F, warning=F, message=F, fig.align='left', fig.height=4, fig.width=4}
# Leave out patients
loo_xval = matrix(ncol=3, nrow=0, dimnames=list(c(), c('net.similarity', 'sum.similarity', 'clusters')))
for (i in 1:nrow(sum.patient.data)) {
  simL = sum.patient.data[-i,c(2:4)]
  rownames(simL) = sum.patient.data[-i,1]

@@

#### Single samples

```{r xval_smp, echo=F, warning=F, message=F, fig.align='left', fig.height=4, fig.width=4}
# Leave out samples?
looS_xval = matrix(ncol=3, nrow=0, dimnames=list(c(), c('net.similarity', 'sum.similarity', 'clusters')))
for (i in 1:20) { 
  simX = sum.patient.data[,c('Status','years')]
  rownames(simX) = sum.patient.data[,'Patient']

@@

## No HGD samples

There is still a difference between the progressors and non, though the p-value does increase. Some of the progressors do start looking more like NPs in their similarity measure but not all of them.

```{r nofinal, echo=F, message=F, warning=F, fig.height=5, fig.width=5}

@@

for (pt in sum.patient.data$Patient) {
  x1 = apclust.data(patient.data[[pt]]$seg.vals, patient.data[[pt]]$info$Samplename)
  if (subset(sum.patient.data, Patient == pt)$Status == 'P' & length(which(grepl('HGD|IMC', patient.data[[pt]]$info$Pathology))) > 0) {
    final = grep('HGD|IMC', patient.data[[pt]]$info$Pathology)
    x1 = apclust.data(patient.data[[pt]]$seg.vals, patient.data[[pt]]$info$Samplename[-final])
  }
  ac = apcluster(negDistMat(x1, r=2), q=q)

@@

fit = glm(Status~clusters, simX, family=binomial(link='logit') )

@@

There do appear to be some patients that are more similar to non-progressors without their HGD samples:
`r pander(subset(simX, Status == 'P' & clusters <= mean(subset(simX, Status == 'NP')$clusters)), caption="These progressor patients fall below the mean number of clusters in the non-progressor group, suggesting they may be more similar to NP before their HGD samples.")`

## AP cluster parameter selection

Just a quick note on the selection of the q parameter for apcluster. The q parameter initializes the exemplar preferences.  q=0.5 starts at the median, values between 0-1 use the sample quantile with threshold q (per documentation).

```{r whyq, echo=F, message=F, warning=F}
sum.patient.data = arrange(sum.patient.data, Status)

pvals = as.data.frame(matrix(ncol=3,nrow=0,dimnames=list(c(), c('p.clusters','mean.clusters', 'sd.clusters'))))
status.clusters = list()
for (qv in seq(0,0.9,0.1)) {
  ## Using the segmentation values
  simPC = sum.patient.data[,c(2:4)]
  rownames(simPC) = sum.patient.data[,1]
  for (pt in sum.patient.data$Patient) {
    patient.data[[pt]]$info = arrange(patient.data[[pt]]$info, Endoscopy.Year, desc(Pathology))

@@
